Proving Logical Atomicity using Lock Invariants\thanks{Supported by the National Science Foundation via grant \#1811894.}

### 2022-05-16 concurrency/ASL 2022/intro.tex
dealt with some of Lennart's comments

Intuitively, a critical section in a lock-based implementation serves the same role as an atomic instruction in a lock-free data structure, and it is not hard to identify linearization points in lock-based implementations---the effects of an update become visible when the lock protecting the updated data is released. The translation of this intuition into proof, however, depends on the specifications used for lock operations. The CSL literature contains two distinct specifications for lock acquire and release:
\begin{enumerate}
\item In Gotsman/Hobor-style specs~\cite{gotsman,oraclesematic}, each lock is associated with a lock invariant $R$, which is gained by \texttt{acquire} operations and restored by \texttt{release} operations.
\item In TaDA-style specs~\cite{tada}, \texttt{acquire} atomically moves a lock from the unlocked state to the locked state, and \texttt{release} does the reverse.
\end{enumerate}
Style 1 can be derived from style 2, and is also a natural fit for atomicity proofs of larger data structures, so it has been the style of choice for most recent verifications~\cite{tada-live,templates}. However, unpublished work by Zhang~\cite{atomic-syncer} shows that style 1 can also be used to prove atomic specifications for data structures.

In this paper, we adapt this approach to the Verified Software Toolchain~\cite{plfcc}, a system for proving correctness of C programs that relies on style 1 lock specs for its soundness proof w.r.t. the CompCert compiler~\cite{compcert}, and has recently been extended with support for general ghost state and atomicity proofs. We demonstrate that 1) \textbf{our approach can be used to prove atomic specs for interesting data structures with fine-grained locking}, and 2) \textbf{these proofs require some additional complications that can be avoided with style 2 lock specs}. The key difference is that in style 1 a thread must gain ownership of the lock invariant assertion for a component before acquiring the lock and reading/modifying the component, while in style 2 the locks and components alike can be considered part of a single abstract state that is accessed at each lock access. However, the top-level specifications proved for the data structure are not affected by the complexity of the lock specs, so proofs of clients of the data structure are no more complex than they would be otherwise. To our knowledge, this is the first formal comparison of the two styles of lock specifications, and our technique should be useful in tools like VST where the older style is deeply integrated into the tool.

@@

>>>>

Intuitively, a critical section in a lock-based implementation serves the same role as an atomic instruction in a lock-free data structure, and it is not hard to identify linearization points in lock-based implementations---the effects of an update become visible when the lock protecting the updated data is released. The translation of this intuition into proof, however, depends on the specifications used for lock operations. The CSL literature contains two distinct specifications for lock acquire and release:
\begin{enumerate}
\item In Gotsman/Hobor-style specs~\cite{gotsman,oraclesematic}, each lock is associated with a lock invariant $R$, which is gained by \texttt{acquire} operations and restored by \texttt{release} operations.
\item In TaDA-style specs~\cite{tada}, \texttt{acquire} atomically moves a lock from the unlocked state to the locked state, and \texttt{release} does the reverse.
\end{enumerate}
Style 2 specs are known to imply style 1 specs, and have been the style of choice for most recent verifications~\cite{tada-live,templates}. However, unpublished work by Zhang~\cite{atomic-syncer} shows that style 1 can also be used to prove atomic specifications for data structures.

In this paper, we adapt this \LB{This reads like ``we carry out Zhang's approach in VST''}\wm{That's basically true.} approach to the Verified Software Toolchain~\cite{plfcc}, a system for proving correctness of C programs that has recently been extended with support for general ghost state and atomicity proofs. The soundness proof for VST~\cite{cpm} lifts the single-threaded correctness theorem of the CompCert compiler~\cite{compcert} to a concurrent setting, and relies explicitly on style 1 lock specs; modifying this proof to rely on atomic operations instead is both theoretically and practically daunting. We demonstrate that 1) \textbf{our approach can be used to prove atomic specs for interesting data structures with fine-grained locking, starting from style 1 lock specs}, and 2) \textbf{these proofs require some additional complications that can be avoided with style 2 lock specs}. The key difference is that in style 1 a thread must gain ownership of a handle to a component's lock before acquiring the lock and reading/modifying the component, while in style 2 the locks and components alike can be considered part of a single abstract state that is accessed at each lock access. However, the top-level specifications proved for the data structure are not affected by the complexity of the lock specs, so proofs of clients of the data structure are no more complex than they would be otherwise.

To our knowledge, this is the first formal comparison of the two styles of lock specifications, and our technique should be useful in tools like VST where the older style is deeply integrated into the tool. This is also the first \emph{foundational} verification of a C implementation against logically atomic specifications; prior tools such as VeriFast~\cite{verifast-conc} have been used to verify C implementations, but are not proved sound against a C semantics and so can avoid the complexities of lock specs w.r.t. concurrent soundness.

@@

\LB{In contrast to all previous wroks, we our work is on full C implementations of binary search trees rather than pseudo code. This is enabled by recently added incorporation of (slightly limited) Iris-style ghost state and invariants (and Iris prof mode), and of TaDa-style atomic specs in VST (citation to William's technical report?)}\wm{It's currently a rejected ITP paper rather than a technical report, sadly.}
